---
title: "V2_generate_analysis_inputs"
author: "Jared G. Young"
date: "2026-08-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prepare input files
This script is to prepare input files for SNP-caller and benchmarking. Cheifly, this will be to calculate true-positive (theoretical) results and compare it against the actual SNP-calling approach. 

As inputs, I will have several files: 

- TP_snp_positions = the accessions and injected SNP positions that will make up the theoretical True Positive SNPs to be detected at different levels. 

- Injected ARGs = a csv file containing the ARGs that were injected into each input file. As such, only a subset of true-positive SNPs should be found in each mock sample. 

What I want: a matrix of the true-positive SNPs (binary, where 0 = not inserted and 1 = inserted) that can be further parsed to calculate how many of the true positive SNPs were actually captured. 

# Write preparation scripts
## One-based data
Here, I am converting the True_positive SNP file into an annotation structure: 
### Reorder for gene/snp set 1: 
```{r}
# load data 
tpsnp_key <- read.csv("v2_tp_snp_data/set_01_one_based_SNP_key.csv")
# load packages
library(tidyr)
library(dplyr)
library(stringr)
# copy header line to split
tpsnp_key$headtosplit <- tpsnp_key$header

# Split_header
tpsnp_key_split <- tpsnp_key |> 
  separate_wider_delim(headtosplit, delim = "|", names = c("acc","type","class","mechanism","group"))
tpsnp_key_split$acc_dup <- tpsnp_key_split$acc
tpsnp_key_split$group_dup <- tpsnp_key_split$group
# concat group-acc
tpsnp_key_split <- tpsnp_key_split |> mutate(acc_dup = str_remove_all(acc_dup, "MEG_"))
tpsnp_key_split <- tpsnp_key_split |>
  unite("grpacc",group_dup,acc_dup, sep = "-")
tpsnp_key_split <- tpsnp_key_split |>
  unite("snp_ann",grpacc,reference_nucleotide,one_based_position,variant_nucleotide, sep = "_")
# Re_concat columns to create SNP headers: 
snp_heads <- tpsnp_key_split |>
  unite("snp_header",acc,type,class,mechanism,group,acc,snp_ann, sep = "|", remove = FALSE)

# Re-order columns to make a header file 
snp_heads <- snp_heads |> relocate(acc, .after = group)
snp_heads <- snp_heads |> relocate(snp_ann, .after = acc)

set1_snps <- snp_heads
# save_rds
#saveRDS(snp_heads, file = "V2_benchmark_data/set1_v4_snp_annotations.rds") # done 8/3/2026
# save csv
# write.csv(snp_heads, "V2_benchmark_data/set1_v4_snp_annotations.csv", row.names = FALSE) # done 8/3/2026
```

### Make yes/no matrix for set1 input fasta files
```{r}
set1_genes <- read.csv("v2_tp_snp_data/set_01_annotations.csv")
R01_genes <- read.csv("v2_tp_snp_data/R01_set1_gene_coord.csv")
# Make new data frame that has a column for each repeat
## R01
### Genes
R01_anns <- as.list(R01_genes$inserted_gene_id)
set1_genes <- set1_genes |> 
  mutate(R01 = as.integer(header %in% R01_anns))
### SNPs 
set1_snps <- set1_snps |>
  mutate(R01 = as.integer(header %in% R01_anns))
rm(R01_anns,R01_genes)

## R02 
R02_genes <- read.csv("v2_tp_snp_data/R02_set1_gene_coord.csv")
### Genes
R02_anns <- as.list(R02_genes$inserted_gene_id)
set1_genes <- set1_genes |> 
  mutate(R02 = as.integer(header %in% R02_anns))
### SNPs 
set1_snps <- set1_snps |>
  mutate(R02 = as.integer(header %in% R02_anns))
rm(R02_anns,R02_genes)

## R03 
R03_genes <- read.csv("v2_tp_snp_data/R03_set1_gene_coord.csv")
### Genes
R03_anns <- as.list(R03_genes$inserted_gene_id)
set1_genes <- set1_genes |> 
  mutate(R03 = as.integer(header %in% R03_anns))
### SNPs 
set1_snps <- set1_snps |>
  mutate(R03 = as.integer(header %in% R03_anns))
rm(R03_anns,R03_genes)

## R04 
R04_genes <- read.csv("v2_tp_snp_data/R04_set1_gene_coord.csv")
### Genes
R04_anns <- as.list(R04_genes$inserted_gene_id)
set1_genes <- set1_genes |> 
  mutate(R04 = as.integer(header %in% R04_anns))
### SNPs 
set1_snps <- set1_snps |>
  mutate(R04 = as.integer(header %in% R04_anns))
rm(R04_anns,R04_genes)

## R05 
R05_genes <- read.csv("v2_tp_snp_data/R05_set1_gene_coord.csv")
### Genes
R05_anns <- as.list(R05_genes$inserted_gene_id)
set1_genes <- set1_genes |> 
  mutate(R05 = as.integer(header %in% R05_anns))
### SNPs 
set1_snps <- set1_snps |>
  mutate(R05 = as.integer(header %in% R05_anns))
rm(R05_anns,R05_genes)

# save_rds
#saveRDS(set1_genes, file = "V2_benchmark_data/set1_v4_R_gene_pa.rds") # done 8/3/2026
#saveRDS(set1_snps, file = "V2_benchmark_data/set1_v4_R_snp_onebased_pa.rds") # done 8/3/2026
# save csv
#write.csv(set1_genes, "V2_benchmark_data/set1_v4_R_gene_pa.csv", row.names = FALSE) # done 8/3/2026
#write.csv(set1_snps, "V2_benchmark_data/set1_v4_R_snp_onebased_pa.csv", row.names = FALSE) # done 8/3/2026
```

### Repeat for set2 genes/snps
```{r}
# load data 
tpsnp_key <- read.csv("v2_tp_snp_data/set_02_one_based_SNP_key.csv")
# copy header line to split
tpsnp_key$headtosplit <- tpsnp_key$header

# Split_header
tpsnp_key_split <- tpsnp_key |> 
  separate_wider_delim(headtosplit, delim = "|", names = c("acc","type","class","mechanism","group"))
tpsnp_key_split$acc_dup <- tpsnp_key_split$acc
tpsnp_key_split$group_dup <- tpsnp_key_split$group
# concat group-acc
tpsnp_key_split <- tpsnp_key_split |> mutate(acc_dup = str_remove_all(acc_dup, "MEG_"))
tpsnp_key_split <- tpsnp_key_split |>
  unite("grpacc",group_dup,acc_dup, sep = "-")
tpsnp_key_split <- tpsnp_key_split |>
  unite("snp_ann",grpacc,reference_nucleotide,one_based_position,variant_nucleotide, sep = "_")
# Re_concat columns to create SNP headers: 
snp_heads <- tpsnp_key_split |>
  unite("snp_header",acc,type,class,mechanism,group,acc,snp_ann, sep = "|", remove = FALSE)

# Re-order columns to make a header file 
snp_heads <- snp_heads |> relocate(acc, .after = group)
snp_heads <- snp_heads |> relocate(snp_ann, .after = acc)

set2_snps <- snp_heads
# save_rds
#saveRDS(snp_heads, file = "V2_benchmark_data/set2_v4_snp_annotations_onebased.rds") # done 8/3/2026
# save csv
#write.csv(snp_heads, "V2_benchmark_data/set2_v4_snp_annotations_onebased.csv", row.names = FALSE) # done 8/3/2026
```

### Make yes/no matrix for set2 input fasta files
```{r}
set2_genes <- read.csv("v2_tp_snp_data/set_02_annotations.csv")
R06_genes <- read.csv("v2_tp_snp_data/R06_set2_gene_coords.csv")
# Make new data frame that has a column for each repeat
## R06
### Genes
R06_anns <- as.list(R06_genes$inserted_gene_id)
set2_genes <- set2_genes |> 
  mutate(R06 = as.integer(header %in% R06_anns))
### SNPs 
set2_snps <- set2_snps |>
  mutate(R06 = as.integer(header %in% R06_anns))
rm(R06_anns,R06_genes)

## R07 
R07_genes <- read.csv("v2_tp_snp_data/R07_set2_gene_coords.csv")
### Genes
R07_anns <- as.list(R07_genes$inserted_gene_id)
set2_genes <- set2_genes |> 
  mutate(R07 = as.integer(header %in% R07_anns))
### SNPs 
set2_snps <- set2_snps |>
  mutate(R07 = as.integer(header %in% R07_anns))
rm(R07_anns,R07_genes)

## R08 
R08_genes <- read.csv("v2_tp_snp_data/R08_set2_gene_coords.csv")
### Genes
R08_anns <- as.list(R08_genes$inserted_gene_id)
set2_genes <- set2_genes |> 
  mutate(R08 = as.integer(header %in% R08_anns))
### SNPs 
set2_snps <- set2_snps |>
  mutate(R08 = as.integer(header %in% R08_anns))
rm(R08_anns,R08_genes)

## R09 
R09_genes <- read.csv("v2_tp_snp_data/R09_set2_gene_coords.csv")
### Genes
R09_anns <- as.list(R09_genes$inserted_gene_id)
set2_genes <- set2_genes |> 
  mutate(R09 = as.integer(header %in% R09_anns))
### SNPs 
set2_snps <- set2_snps |>
  mutate(R09 = as.integer(header %in% R09_anns))
rm(R09_anns,R09_genes)

## R10 
R10_genes <- read.csv("v2_tp_snp_data/R10_set2_gene_coords.csv")
### Genes
R10_anns <- as.list(R10_genes$inserted_gene_id)
set2_genes <- set2_genes |> 
  mutate(R10 = as.integer(header %in% R10_anns))
### SNPs 
set2_snps <- set2_snps |>
  mutate(R10 = as.integer(header %in% R10_anns))
rm(R10_anns,R10_genes)

# save_rds
#saveRDS(set2_genes, file = "V2_benchmark_data/set2_v4_R_gene_pa.rds") # done 8/3/2026
#saveRDS(set2_snps, file = "V2_benchmark_data/set2_v4_R_snp_onebased_pa.rds") # done 8/3/2026
# save csv
#write.csv(set2_genes, "V2_benchmark_data/set2_v4_R_gene_pa.csv", row.names = FALSE) # done 8/3/2026
#write.csv(set2_snps, "V2_benchmark_data/set2_v4_R_snp_onebased_pa.csv", row.names = FALSE) # done 8/3/2026
```

## Repeat for 0-based data
#### Reorder for gene/snp set 1: 
```{r}
# load data 
tpsnp_key <- read.csv("v2_tp_snp_dataset_01_zero_based_SNP_key.csv")

# copy header line to split
tpsnp_key$headtosplit <- tpsnp_key$header

# Split_header
tpsnp_key_split <- tpsnp_key |> 
  separate_wider_delim(headtosplit, delim = "|", names = c("acc","type","class","mechanism","group"))
tpsnp_key_split$acc_dup <- tpsnp_key_split$acc
tpsnp_key_split$group_dup <- tpsnp_key_split$group
# concat group-acc
tpsnp_key_split <- tpsnp_key_split |> mutate(acc_dup = str_remove_all(acc_dup, "MEG_"))
tpsnp_key_split <- tpsnp_key_split |>
  unite("grpacc",group_dup,acc_dup, sep = "-")
tpsnp_key_split <- tpsnp_key_split |>
  unite("snp_ann",grpacc,reference_nucleotide,zero_based_position,variant_nucleotide, sep = "_")
# Re_concat columns to create SNP headers: 
snp_heads <- tpsnp_key_split |>
  unite("snp_header",acc,type,class,mechanism,group,acc,snp_ann, sep = "|", remove = FALSE)

# Re-order columns to make a header file 
snp_heads <- snp_heads |> relocate(acc, .after = group)
snp_heads <- snp_heads |> relocate(snp_ann, .after = acc)

set1_snps <- snp_heads
# save_rds
##saveRDS(snp_heads, file = "zb_benchmark_data/set1_v4_snp_annotations_zerobased.rds") # done 7/24/2026
# save csv
##write.csv(snp_heads, "zb_benchmark_data/set1_v4_snp_annotations_zerobased.csv", row.names = FALSE) # done 7/24/2026
```

### Make yes/no matrix for set1 input fasta files
```{r}
set1_genes <- read.csv("v2_tp_snp_dataset_01_arg_annotations.csv")
R01_genes <- read.csv("v2_tp_snp_data/R01_set1_gene_coords.csv")
# Make new data frame that has a column for each repeat
## R01
### Genes
R01_anns <- as.list(R01_genes$inserted_gene_id)
set1_genes <- set1_genes |> 
  mutate(R01 = as.integer(header %in% R01_anns))
### SNPs 
set1_snps <- set1_snps |>
  mutate(R01 = as.integer(header %in% R01_anns))
rm(R01_anns,R01_genes)

## R02 
R02_genes <- read.csv("v2_tp_snp_data/R02_set1_gene_coord.csv")
### Genes
R02_anns <- as.list(R02_genes$inserted_gene_id)
set1_genes <- set1_genes |> 
  mutate(R02 = as.integer(header %in% R02_anns))
### SNPs 
set1_snps <- set1_snps |>
  mutate(R02 = as.integer(header %in% R02_anns))
rm(R02_anns,R02_genes)

## R03 
R03_genes <- read.csv("v2_tp_snp_data/R03_set1_gene_coord.csv")
### Genes
R03_anns <- as.list(R03_genes$inserted_gene_id)
set1_genes <- set1_genes |> 
  mutate(R03 = as.integer(header %in% R03_anns))
### SNPs 
set1_snps <- set1_snps |>
  mutate(R03 = as.integer(header %in% R03_anns))
rm(R03_anns,R03_genes)

## R04 
R04_genes <- read.csv("v2_tp_snp_data/R04_set1_gene_coord.csv")
### Genes
R04_anns <- as.list(R04_genes$inserted_gene_id)
set1_genes <- set1_genes |> 
  mutate(R04 = as.integer(header %in% R04_anns))
### SNPs 
set1_snps <- set1_snps |>
  mutate(R04 = as.integer(header %in% R04_anns))
rm(R04_anns,R04_genes)

## R05 
R05_genes <- read.csv("v2_tp_snp_data/R05_set1_gene_coord.csv")
### Genes
R05_anns <- as.list(R05_genes$inserted_gene_id)
set1_genes <- set1_genes |> 
  mutate(R05 = as.integer(header %in% R05_anns))
### SNPs 
set1_snps <- set1_snps |>
  mutate(R05 = as.integer(header %in% R05_anns))
rm(R05_anns,R05_genes)

# save_rds
saveRDS(set1_genes, file = "V2_benchmark_data/set1_v4_R_gene_pa.rds") # done 7/24/2026
saveRDS(set1_snps, file = "V2_benchmark_data/set1_v4_R_snp_zerobased_pa.rds") # done 7/24/2026
# save csv
write.csv(set1_genes, "V2_benchmark_data/set1_v4_R_gene_pa.csv", row.names = FALSE) # done 7/24/2026
write.csv(set1_snps, "V2_benchmark_data/set1_v4_R_snp_zerobased_pa.csv", row.names = FALSE) # done 7/24/2026
```

### Repeat for set2 genes/snps
```{r}
# load data 
tpsnp_key <- read.csv("tp_snp_data/set_02_data/set_02_zero_based_SNP_key.csv")
# copy header line to split
tpsnp_key$headtosplit <- tpsnp_key$header

# Split_header
tpsnp_key_split <- tpsnp_key |> 
  separate_wider_delim(headtosplit, delim = "|", names = c("acc","type","class","mechanism","group"))
tpsnp_key_split$acc_dup <- tpsnp_key_split$acc
tpsnp_key_split$group_dup <- tpsnp_key_split$group
# concat group-acc
tpsnp_key_split <- tpsnp_key_split |> mutate(acc_dup = str_remove_all(acc_dup, "MEG_"))
tpsnp_key_split <- tpsnp_key_split |>
  unite("grpacc",group_dup,acc_dup, sep = "-")
tpsnp_key_split <- tpsnp_key_split |>
  unite("snp_ann",grpacc,reference_nucleotide,zero_based_position,variant_nucleotide, sep = "_")
# Re_concat columns to create SNP headers: 
snp_heads <- tpsnp_key_split |>
  unite("snp_header",acc,type,class,mechanism,group,acc,snp_ann, sep = "|", remove = FALSE)

# Re-order columns to make a header file 
snp_heads <- snp_heads |> relocate(acc, .after = group)
snp_heads <- snp_heads |> relocate(snp_ann, .after = acc)

set2_snps <- snp_heads
# save_rds
##saveRDS(snp_heads, file = "zb_benchmark_data/set2_v4_snp_annotations_zerobased.rds") # done 7/24/2026
# save csv
##write.csv(snp_heads, "zb_benchmark_data/set2_v4_snp_annotations_zerobased.csv", row.names = FALSE) # done 7/24/2026
```

### Make yes/no matrix for set2 input fasta files
```{r}
set2_genes <- read.csv("tp_snp_data/set_02_data/set_02_annotations.csv")
R06_genes <- read.csv("v2_tp_snp_data/R06_set2_gene_coords.csv")
# Make new data frame that has a column for each repeat
## R06
### Genes
R06_anns <- as.list(R06_genes$inserted_gene_id)
set2_genes <- set2_genes |> 
  mutate(R06 = as.integer(header %in% R06_anns))
### SNPs 
set2_snps <- set2_snps |>
  mutate(R06 = as.integer(header %in% R06_anns))
rm(R06_anns,R06_genes)

## R07 
R07_genes <- read.csv("v2_tp_snp_data/R07_set2_gene_coords.csv")
### Genes
R07_anns <- as.list(R07_genes$inserted_gene_id)
set2_genes <- set2_genes |> 
  mutate(R07 = as.integer(header %in% R07_anns))
### SNPs 
set2_snps <- set2_snps |>
  mutate(R07 = as.integer(header %in% R07_anns))
rm(R07_anns,R07_genes)

## R08 
R08_genes <- read.csv("v2_tp_snp_data/R08_set2_gene_coords.csv")
### Genes
R08_anns <- as.list(R08_genes$inserted_gene_id)
set2_genes <- set2_genes |> 
  mutate(R08 = as.integer(header %in% R08_anns))
### SNPs 
set2_snps <- set2_snps |>
  mutate(R08 = as.integer(header %in% R08_anns))
rm(R08_anns,R08_genes)

## R09 
R09_genes <- read.csv("v2_tp_snp_data/R09_set2_gene_coords.csv")
### Genes
R09_anns <- as.list(R09_genes$inserted_gene_id)
set2_genes <- set2_genes |> 
  mutate(R09 = as.integer(header %in% R09_anns))
### SNPs 
set2_snps <- set2_snps |>
  mutate(R09 = as.integer(header %in% R09_anns))
rm(R09_anns,R09_genes)

## R10 
R10_genes <- read.csv("v2_tp_snp_data/R10_set2_gene_coords.csv")
### Genes
R10_anns <- as.list(R10_genes$inserted_gene_id)
set2_genes <- set2_genes |> 
  mutate(R10 = as.integer(header %in% R10_anns))
### SNPs 
set2_snps <- set2_snps |>
  mutate(R10 = as.integer(header %in% R10_anns))
rm(R10_anns,R10_genes)

# save_rds
##saveRDS(set2_genes, file = "zb_benchmark_data/set2_v4_R_gene_pa.rds") # done 7/24/2026
##saveRDS(set2_snps, file = "zb_benchmark_data/set2_v4_R_snp_zerobased_pa.rds") # done 7/24/2026
# save csv
##write.csv(set2_genes, "zb_benchmark_data/set2_v4_R_gene_pa.csv", row.names = FALSE) # done 7/24/2026
##write.csv(set2_snps, "zb_benchmark_data/set2_v4_R_snp_zerobased_pa.csv", row.names = FALSE) # done 7/24/2026
```